Cosmological Hydrodynamic Simulations of Preferential Accretion in the SMBH of Milky Way Size Galaxies

### 2016-06-11 ms.tex
Full Complete Paper Written, Waiting for Kelly Edits

% For thesis: M-sigma plot:
%\begin{figure}
%\centerline{\resizebox{0.75\hsize}{!}{\includegraphics[angle=0]{fancymsigma}}}
%\caption[]{Gas fraction across redshift for galaxy (solid lines) and central BH (dashed lines). Green lines signify gas fractions accreted via mergers and blue lines designate gas accreted via cold flow.}
%\label{msigma} 
%\end{figure}
% Cite: Xiao

@@

% For my thesis: The unshocked flow filaments that feed thses galaxies are dependent on the galaxy's mass. Smoothly accreted gas will fuel through unshocked gas filaments until the galaxy halo reaches a a mass near 10$^{11} M_{\odot}$. Once the galaxy's mass is greater than this threshold, the mass will reach the virial radius of the halo and the rapid increase in temperature within the radius causes the gas to shock as its temperature quickly increases.

@@

Since, there are uncertainties in the formation of black holes ``seeds,'' we implement a BH seeding method that is broadly consistent the theories of direct collapse black holes \citep{Couchman1986, Abel2002, Bromm2004} and Population III stellar remnants\citep{Loeb1994, Eisenstein1995, Koushiappas2004, Begelman2006, Lodato2006}. While this method allows the BH formation process to remain physically motivated. BH seeds form if their parent gas particle match the criteria required for star formation and also maintain zero metallicity. \citep{Stinson2006} A probability of $\chi_{\rm seed}$ ~ 0.01 is applied to determine whether a gas particle (with the above specifications) will become a BH seed with a mass of M$_{\rm BH}$ = 2.28 $\times$ 10$^5$ M$_{\odot}$, the same mass as its parent gas particle. This probability was chosen to match the predicted occupation probability of BH seeds at z $\sim$ 3. \citep{Volonteri2008} 
%This circumvents the Soltan argument wherein mergers alone are not the primary feeders of BHs, allowing for accretion growth to dominate \citep{Soltan1982}.  
BH seed formation ends at z $\sim$ 3.5 due to the dilution of pristine gas by efficient metal diffusion expected after the first population of supernovae. The requirement that BH seeds must form from zero metallicity gas particles also causes BH formation to be constrained in areas of early star formation bursts, where the earliest and most massive halos are expected to form in the simulation. BH formation is dependent only on their local environment, neglecting any large-scale properties of the host halo. Black holes are not fixed within the center of their host, allowing them to be dynamically affected by mergers and other perturbations within the galaxy. Nevertheless, BHs are kept near their host centers by choosing dark matter particles masses in high-resolution regions to be on the same order as gas particle masses which minimizes two-body interactions. \citep[Change citations?]{Bellovary2010,Zolotov2012}

@@

The Amiga Halo Finder identifies all of the halos in the simulations by identifying the virial radii of the halos based on an overdensity using the criterion for a flat universe. \citep{Knebe2001,Knollmann2009,Gill2004} We select the primary halo by determining the most massive galaxy at z=0 and locating the central SMBH. For our Gasoline simulations, both the primary halo in h258 and h277 had final masses on the order of $10^{12} M_{\odot}$ and a formation redshift of z~4.

@@

The galaxy, h258, is characterized by an active merger history including a major merger at z=1. The merger can be seen affecting the cumulative black hole mass (Figure \ref{h258allmassgas}) as well as the gas fractions within the both the SMBH and galaxy (Figure \ref{h258numfrac}). 
Figure \ref{h258allmassgas} shows h258's central SMBH's cumulative mass as a function of time (lower axis) and redshift (upper axis). The black dashed line indicates the total cumulative BH mass (including both mass from gas and BH mergers), while the black solid line indicates only the total gas mass accreted by the BH. The blue dot-dashed line indicates the gas mass accreted via unshocked gas. The green solid line indicates the gas mass accreted through mergers. The red dashed line indicates gas mass that was shocked upon entry into the halo. Though this figure shows that the primary source of mass gain at high redshifts is through mergers with other black holes, there is a significant increase in the amount of gas adding to the mass of the SMBH starting at z $\sim$ 1, where we know a major merger takes place. \textbf{[Reference merger tree here once you've added it.]} Figure \ref{h258numfrac} shows the fractions of unshocked and clumpy gas compared to all accreted gas as a function of redshift. It confirms that unshocked gas (blue) makes up the majority of gas entering the galaxy (solid lines) at early times; nevertheless, around the time of the merger (z $\sim$ 1) a marked change occurs in the gas fractions within the black hole (dashed lines) where clumpy gas (green) becomes the main component of total gas.

@@

Our quiescent galaxy, h277, was characterized by few mergers in its past, its final major merger occuring around z $\sim$ 3. This merger is visible in Figure \ref{h277allmassgas}, though more subtle than the influx of gas in h258. (Linestyles as in Figure \ref{h258allmassgas}.) In h277, gas seemed to be gained at a more steady rate and at fairly high redshifts comprises a larger portion of the BH's total mass than seen in the previous, more active galaxy. The gas fractions (Figure \ref{h277numfrac}) in h277 (solid lines) and SMBH (dashed lines) are also markedly different than that of a galaxy with an active merger history. While it appears neither clumpy or unshocked gas (green and blue, respectively) particularly dominates at higher redshifts in the galaxy, merger gas becomes prominent in the SMBH at times when mergers occur. \textbf{[Ref merger tree when included.]} Again, this implies some physical process allowing merger gas to readily fuel black hole growth. Though, it is important to note that by z $\sim$ 0, smoothly accreted, unshocked gas again dominates in both the SMBH and host galaxy. As in our more active galaxy, it appears that the role of clumpy gas as it fuels the SMBH becomes important when mergers affect the galaxy; however, with h277 having a significantly less active merger history, unshocked, smooth accretion inevitably dominates the total gas supply.

This result can be more clearly seen in Figure \ref{h277stackfrac}. (Colors as in \ref{h258stackfrac}.) Similar to Bellovary et. al. 2013 study, which found SMBH accretion in high mass, high redshift galaxies mirror their hosts in terms of accreted gas, h277's SMBH accretes a similar fraction of each type of gas, comparable to its host galaxy. Smoothly accreted, unshocked gas represents about half of the total gas in both the SMBH and the galaxy. Clumpy gas also maintains about 30\% of both, while an increase in shocked gas occurs, from 13 \% in the galaxy to 22 \% in the SMBH. %; however, we argue this may be due to supernova contamination in the simulation. 
So by present day, h277 found no strong preference for the SMBH to prefer any one kind of gas in its accretion. Like the SMBHs in the study of \cite{Bellovary2013}, nearly the same fractions of gas entering the halo would then be accreted onto the SMBH (with the exception of shocked gas). Therefore, this may imply that without some physical driver, like the major merger of h258, clumpy gas is unlikely to dominate the gas accreted into the SMBH.

@@

Just as we found in the lower resolution simulation, the high resolution sees a distinct preference for accreting gas that has been gained through the many mergers in its history.

\section{Conclusion}

[Still need to write; Yes still need to write]

@@

\begin{figure}
\centerline{\resizebox{0.75\hsize}{!}{\includegraphics[angle=0]{h258_am_tsnew_228}}}
\caption[]{ Cumulative distribution of angular momentum of the gas particles accreted onto h258 at the time of the major merger (z ~ 1). There are about 95,000 gas particles accreted at this timestep, which gas particles accreted onto the main halo and central black hole distinguished by solid and dashed lines. The green, blue, and red lines indicate clumpy, unshocked, and shocked gas, respectively.}
\label{h258angmom_merger} 
\end{figure}

@@

@@

\begin{figure}
\centerline{\resizebox{0.75\hsize}{!}{\includegraphics[angle=0]{hrh258_am_tsnew_1584}}}
\caption[]{ Cumulative distribution of angular momentum of the gas particles accreted onto the high resolution ChaNGA h258 galaxy at the time of the major merger (z ~ 1). There are about 450,000 gas particles accreted at this timestep, which gas particles accreted onto the main halo and central black hole distinguished by solid and dashed lines. The green, blue, and red lines indicate clumpy, unshocked, and shocked gas, respectively.}
\label{hrh258angmom_merger} 
\end{figure}

\begin{figure}
\centerline{\resizebox{0.75\hsize}{!}{\includegraphics[angle=0]{hrh258_stackbarfractions}}}
\caption[]{Gas fractions of the gas particles accreted by the high resolution ChaNGA h258 by the main halo (left) and the SMBH (right), distinguished by type. Blue, green, and red distinguish gas gained through unshocked gas, gained through mergers, and gas shocked upon entry, respectively. Yellow indicates gas that existed within the main halo upon formation; this ``early'' gas is negligable ($<$ 1 \%) within the SMBH.}
\label{h277stackfrac} 
\end{figure}

>>>>

% For thesis: M-sigma plot:
\begin{figure}
\centerline{\resizebox{0.75\hsize}{!}{\includegraphics[angle=0]{fancymsigma}}}
\caption[]{Gas fraction across redshift for galaxy (solid lines) and central BH (dashed lines). Green lines signify gas fractions accreted via mergers and blue lines designate gas accreted via cold flow.}
\label{msigma} 
\end{figure}
% Cite: Xiao

@@

@@

Since, there are uncertainties in the formation of black holes ``seeds,'' we implement a BH seeding method that is broadly consistent the theories of direct collapse black holes \citep{Couchman1986, Abel2002, Bromm2004} and Population III stellar remnants\citep{Loeb1994, Eisenstein1995, Koushiappas2004, Begelman2006, Lodato2006}. While this method allows the BH formation process to remain physically motivated. BH seeds form if their parent gas particle match the criteria required for star formation and also maintain zero metallicity. \citep{Stinson2006} A probability of $\chi_{\rm seed}$ $\sim$ 0.01 is applied to determine whether a gas particle (with the above specifications) will become a BH seed with a mass of M$_{\rm BH}$ = 2.28 $\times$ 10$^5$ M$_{\odot}$, the same mass as its parent gas particle. This probability was chosen to match the predicted occupation probability of BH seeds at z $\sim$ 3. \citep{Volonteri2008} 
%This circumvents the Soltan argument wherein mergers alone are not the primary feeders of BHs, allowing for accretion growth to dominate \citep{Soltan1982}.  
BH seed formation ends at z $\sim$ 3.5 due to the dilution of pristine gas by efficient metal diffusion expected after the first population of supernovae. The requirement that BH seeds must form from zero metallicity gas particles also causes BH formation to be constrained in areas of early star formation bursts, where the earliest and most massive halos are expected to form in the simulation. BH formation is dependent only on their local environment, neglecting any large-scale properties of the host halo. Black holes are not fixed within the center of their host, allowing them to be dynamically affected by mergers and other perturbations within the galaxy. Nevertheless, BHs are kept near their host centers by choosing dark matter particles masses in high-resolution regions to be on the same order as gas particle masses which minimizes two-body interactions. \citep[Change citations?]{Bellovary2010,Zolotov2012}

@@

The Amiga Halo Finder identifies all of the halos in the simulations by identifying the virial radii of the halos based on an overdensity using the criterion for a flat universe. \citep{Knebe2001,Knollmann2009,Gill2004} We select the primary halo by determining the most massive galaxy at z=0 and locating the central SMBH. For our Gasoline simulations, both the primary halo in h258 and h277 had final masses on the order of $10^{12} M_{\odot}$ and a formation redshift of z$\sim$4.

@@

The galaxy, h258, is characterized by an active merger history including a major merger at z=1 (Figure \ref{h258face}). The merger can be seen affecting the cumulative black hole mass (Figure \ref{h258allmassgas}) as well as the gas fractions within the both the SMBH and galaxy (Figure \ref{h258numfrac}). 
Figure \ref{h258allmassgas} shows h258's central SMBH's cumulative mass as a function of time (lower axis) and redshift (upper axis). The black dashed line indicates the total cumulative BH mass (including both mass from gas and BH mergers), while the black solid line indicates only the total gas mass accreted by the BH. The blue dot-dashed line indicates the gas mass accreted via unshocked gas. The green solid line indicates the gas mass accreted through mergers. The red dashed line indicates gas mass that was shocked upon entry into the halo. Though this figure shows that the primary source of mass gain at high redshifts is through mergers with other black holes, there is a significant increase in the amount of gas adding to the mass of the SMBH starting at z $\sim$ 1, where we know a major merger takes place. \textbf{[Reference merger tree here once you've added it.]} Figure \ref{h258numfrac} shows the fractions of unshocked and clumpy gas compared to all accreted gas as a function of redshift. It confirms that unshocked gas (blue) makes up the majority of gas entering the galaxy (solid lines) at early times; nevertheless, around the time of the merger (z $\sim$ 1) a marked change occurs in the gas fractions within the black hole (dashed lines) where clumpy gas (green) becomes the main component of total gas.

@@

Our quiescent galaxy, h277, was characterized by few mergers in its past, its final major merger occuring around z $\sim$ 3 (Figure \ref{h277face}). This merger is visible in Figure \ref{h277allmassgas}, though more subtle than the influx of gas in h258. (Linestyles as in Figure \ref{h258allmassgas}.) In h277, gas seemed to be gained at a more steady rate and at fairly high redshifts comprises a larger portion of the BH's total mass than seen in the previous, more active galaxy. The gas fractions (Figure \ref{h277numfrac}) in h277 (solid lines) and SMBH (dashed lines) are also markedly different than that of a galaxy with an active merger history. While it appears neither clumpy or unshocked gas (green and blue, respectively) particularly dominates at higher redshifts in the galaxy, merger gas becomes prominent in the SMBH at times when mergers occur. \textbf{[Ref merger tree when included.]} Again, this implies some physical process allowing merger gas to readily fuel black hole growth. Though, it is important to note that by z $\sim$ 0, smoothly accreted, unshocked gas again dominates in both the SMBH and host galaxy. As in our more active galaxy, it appears that the role of clumpy gas as it fuels the SMBH becomes important when mergers affect the galaxy; however, with h277 having a significantly less active merger history, unshocked, smooth accretion inevitably dominates the total gas supply.

This result can be more clearly seen in Figure \ref{h277stackfrac}. (Colors as in \ref{h258stackfrac}.) Similar to Bellovary et. al. 2013 study, which found SMBH accretion in high mass, high redshift galaxies mirror their hosts in terms of accreted gas, h277's SMBH accretes a similar fraction of each type of gas, comparable to its host galaxy. Smoothly accreted, unshocked gas represents about half of the total gas in both the SMBH and the galaxy. Clumpy gas also maintains about 30\% of both, while an increase in shocked gas occurs, from 13\% in the galaxy to 22\% in the SMBH. %; however, we argue this may be due to supernova contamination in the simulation. 
So by present day, h277 found no strong preference for the SMBH to prefer any one kind of gas in its accretion. Like the SMBHs in the study of \cite{Bellovary2013}, nearly the same fractions of gas entering the halo would then be accreted onto the SMBH (with the exception of shocked gas). Therefore, this may imply that without some physical driver, like the major merger of h258, clumpy gas is unlikely to dominate the gas accreted into the SMBH.

@@

Just as we found in the same galaxy, h258, of the lower resolution simulation, the high resolution h258 galaxy sees a distinct preference for accreting gas that has been gained through the many mergers in its history. In this version, h258 is characterized by two major mergers; the first occurs at z $\sim$ 1.8 and the second at z $\sim$ 1.2. Unlike its predecessor, this h258 has a smoother history of overall gas accretion (Figure \ref{hrh258allmassgas}). (Linestyles as in Figure \ref{h258allmassgas}.) However, the transitions between when smooth, unshocked accretion and mergers dominate is clearly distinguished. While clumpy gas (green) dominates at the earliest time, unshocked gas (blue) overtakes it for a short time ($\sim$ 2 Gyr) before clumpy gas once again dominates by z $\sim$ 1.5.

This low redshift transition to a clumpy gas preference results in the large fraction of clumpy gas seen in the SMBH (Figure \ref{hrh258stackfrac}). (Colors as in Figure \ref{h258stackfrac}.) We can see from the figure that comparable to the Gasoline h258 version, the amount of unshocked, cold gas in the SMBH (24\%) is less than that entering the galaxy (36\%). The shocked gas, however, sees a sharper decline in the amount of shocked gas that enters the SMBH ($\sim$2\%) than in the host (8\%). This differs from the Gasoline h258 results and may be attributed to the higher resolution of the second iteration. [QUESTION FOR JILLIAN: Is this true? I was wondering if the higher SN Feedback in ChaNGa might cause more problems through misclassified shocked gas, but I'm thinking that the higher resolution may fix this issue.] Figure \ref{hrh258stackfrac} also explicitely shows how the fraction of clumpy gas accreted by the SMBH (74\%) is significantly larger than the fraction of clumpy gas accreted by the host galaxy (56\%).  \textbf{This result, which repeats the analysis with an improved and higher resolution simulation, confirms the initial study's findings!} This confirmation makes us confident that these results are well founded.

In keeping with the analysis of the Gasoline galaxies, we followed up the study of the galaxy and SMBH gas fractions by examining the angular momentum of the accreted gas particles. Figure \ref{hrh258angmom} (Colors as in Figure \ref{h258angmom}.) shows a cumulative distribution of the gas particles' angular momentum as they enter the halo, distinguished between gas that ends up in the SMBH (dashed line) or the galaxy (solid lines) at z=0. This plot is consistent with Figure \ref{h258angmom}, as the angular momentum of gas entering the SMBH is lower when comparing between types of gas. In the ChaNGA h258 case, however, we see that the lowest angular momentum gas is comprised of both clumpy and unshocked gas, unlike in the Gasoline version which clearly showed that unshocked gas entered the SMBH with the lowest angular momentum. We attribute this variation to the subtle difference between the histories of the two simulations, including the two mergers which characterize the higher resolution version. This can be seen if we examine Figure \ref{hrh258angmom_merger2} which shows the cumulative distribution of the angular momentum of the incoming gas particles at the time of the second merger. (Colors and linestyles as in Figure \ref{h258angmom_merger}.) Figure \ref{hrh258angmom_merger2} explicitly shows that the gas ending up in the SMBH enters with the lowest angular momentum. This result is similar to what we found in the Gasoline h258 simulation, however, in the previous case it appeared the all the low angular momentum gas entering at the merger timestep ended up in the galaxy. Yet the distinction remains as there appears to be an influx of gas with low angular momentum feeding the black hole at this timestep that is not associated with the merger occurring at z $\sim$ 1.2. This may be attributed to the prior merger at z $\sim$ 1.8 (Figure \ref{hrh258angmom_merger1}) during and after which a large influx of low angular momentum gas enters the galaxy and SMBH. (Color and linestyles as in Figure \ref{hrh258angmom_merger2}) [Talk to Jillian and Kelly about this. Is this a reasonable explanation?]

\section{Conclusion}
This study examines the gas accretion onto two fully cosmological simulations of Milky Way-size galaxies to redshift z $=$ 0, one with a quiescent merger history and one with major mergers characterizing its past. A second iteration of the latter, simulated with higher resolution and improved physics, is also examined. In each case, we trace the gas into the SMBHs at their centers and differentiate the gas accreted onto the galaxy and SMBHs by origin. Gas gained through mergers is classified as ``clumpy'' gas and smoothly accreted gas is seperated into ``shocked'' and ``unshocked'' categories. Our goal is to determine what types of gas are primarily feeding the SMBH and the galaxies of this class, and to determine what effects the merger histories of the galaxies may have on these processes.

When we examine the accretion of the galaxy with a quiescent history, we note that smooth accretion is the dominant feeding mechanism for both the galaxy and its central SMBH. We find that the fractions of each type of gas comprising the overall mass of the SMBH are comparable to the fractions of the gases making up its host. This result is consistent with the results of \cite{Bellovary2013} which similarly analyized high mass, high redshift galaxies and found the gas composition of the SMBHs mirror their host.

Contrary to these previous results, when we examined the galaxy with an active merger history, we determined that the SMBH at the center more readily accretes gas gained through mergers. \textbf{In both the low and high resolution cases, we see a significant increase in the clumpy gas accreted by the SMBH compared to its host.}

The angular momentum of the accreted gas as it enters the galaxy sheds some light on the mechanism driving this preferentially accreted clumpy gas. Smoothly accreted gas, which enters with galaxy with a wide range of angular momentum, may become additionally torqued by the already existing disk. Meanwhile, gas entering through mergers is restricted by the direction of its entry and allows for a small range of low angular momentum gas to enter the main halo. This gives clumpy gas the advantage of falling more readily to the center and accreting onto the SMBH. Considering all origins of gas, we see a clear distinction wherein lower angular momentum gas preferentially feeds the SMBH.

This however, does not remain true when we look again at the galaxy with a quiescent history. In this case, some of the gas entering the galaxy with the lowest angular momentum remains outside of the SMBH. This results implies that some of the physical processes affecting the galaxy with an active merger history accounts for the ready funneling of low angular momentum to the central region of the black hole.

While the examination of these two extreme cases of galaxy merger history depict a class of galaxy with varying SMBH accretion methods, a further study of intermediate cases is required to begin understanding the broad spectrum of Milky Way-mass galaxy accretion. It is clear through this study that the presence (or absence) of major mergers can play an important role in the final compositions of central SMBHs, but the question of how important these mergers are remains to be seen.

@@

\begin{figure}
\centerline{\resizebox{0.75\hsize}{!}{\includegraphics[angle=0]{h258_am_tsnew_228}}}
\caption[]{ Cumulative distribution of angular momentum of the gas particles accreted onto h258 at the time of the major merger (z $\sim$ 1). There are about 95,000 gas particles accreted at this timestep, which gas particles accreted onto the main halo and central black hole distinguished by solid and dashed lines. The green, blue, and red lines indicate clumpy, unshocked, and shocked gas, respectively.}
\label{h258angmom_merger} 
\end{figure}

@@

\begin{figure}
\centerline{\resizebox{0.75\hsize}{!}{\includegraphics[angle=0]{hrh258_stackbarfractions}}}
\caption[]{Gas fractions of the gas particles accreted by the high resolution ChaNGA h258 by the main halo (left) and the SMBH (right), distinguished by type. Blue, green, and red distinguish gas gained through unshocked gas, gained through mergers, and gas shocked upon entry, respectively. Yellow indicates gas that existed within the main halo upon formation; this ``early'' gas is negligable ($<$ 1 \%) within the SMBH.}
\label{hrh258stackfrac} 
\end{figure}

@@

\begin{figure}
\centerline{\resizebox{0.75\hsize}{!}{\includegraphics[angle=0]{hrh258_am_tsnew_1584}}}
\caption[]{ Cumulative distribution of angular momentum of the gas particles accreted onto the high resolution ChaNGA h258 galaxy at the time of the second major merger (z $\sim$ 1.2). There are about 450,000 gas particles accreted at this timestep, which gas particles accreted onto the main halo and central black hole are distinguished by solid and dashed lines. The green, blue, and red lines indicate clumpy, unshocked, and shocked gas, respectively.}
\label{hrh258angmom_merger2} 
\end{figure}

\begin{figure}
\centerline{\resizebox{0.75\hsize}{!}{\includegraphics[angle=0]{hrh258_am_tsnew_1128}}}
\caption[]{ Cumulative distribution of angular momentum of the gas particles accreted onto the high resolution ChaNGA h258 galaxy at the time of the first major merger (z $\sim$ 1.8). There are about 900,000 gas particles accreted at this timestep, which gas particles accreted onto the main halo and central black hole are distinguished by solid and dashed lines. The green, blue, and red lines indicate clumpy, unshocked, and shocked gas, respectively.}
\label{hrh258angmom_merger1} 
\end{figure}
